Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-08-28 paper/manuscript.tex
Mainly responses to review of uncertainty quantification updates by @apbhati1. Also includes updates to authorship.

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[11]{Agastya P. Bhati}
\author[18]{Stefan Boresch}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[11, 12, 13]{Peter V. Coveney}
\author[15]{Jonathan W. Essex}
\author[19]{Emilio Gallicchio}
\author[9]{Vytautas Gapsys}
\author[9]{David F. Hahn}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[15,16]{William G. Poole}
\author[6]{Samarjeet Prasad}
\author[2,7]{Andrea Rizzi}
\author[17]{Jenke Scheen}
\author[14]{David R. Slochower}
\author[15]{Oliver J. Melling}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
\affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}
\affil[15]{School of Chemistry and Chemical Engineering, University of Southampton, Southampton, SO17 1BJ, United Kingdom}
\affil[16]{Astex Pharmaceuticals, 436 Cambridge Science Park, Milton Road, Cambridge, CB4 0QA, United Kingdom}
\affil[17]{Open Molecular Software Foundation, Davis, CA 95618, USA}
\affil[18]{University of Vienna, Faculty of Chemistry, Institute of Computational Biological Chemistry, Wahringer Str. 17, A-1090 Vienna}
\affil[19]{Department of Chemistry and Biochemistry, Brooklyn College of the City University of New York, New York, NY, 11210
Ph.D. Program in Chemistry, The Graduate Center of the City University of New York, New York, NY, 10016
Ph.D. Program in Biochemistry, The Graduate Center of the City University of New York, New York, NY, 10016}

%
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Stefan Boresch}{0000-0002-2793-6656}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Emilio Gallicchio}{0000-0002-2606-4913}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
\orcid{Peter V. Coveney}{0000-0002-8787-7256}
\orcid{William G. Poole}{0009-0003-2441-8794}
\orcid{Oliver J. Melling}{0000-0001-6243-7433}
\orcid{Jonathan W. Essex}{0000-0003-2639-2746}

@@

%There is growing literature describing the theoretical underpinning behind such intrinsic non-reproducibility of MD simulation based predictions. Coveney \textit{et al. }~\cite{coveney2016} have noted that, for a complex dynamical system to reach equilibrium, the dynamics must at least be ``mixing'' which is an established result in ergodic theory. A system which is ``mixing'' is also ``ergodic'' by definition but the reverse is not true. This implies that MD trajectories are chaotic, in other words they are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible. Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}. In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e. statistically robust) free energy predictions. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017} which is based on the use of ensemble methods. It produces excellent results for large scale benchmarking datasets for both RBFEs~\cite{bhati2022} and ABFEs~\cite{bhati2025}. It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision. Further, this method is extremely flexible and adaptive. For example, one can simply add additional ensemble members as necessary or desired, without having to re-run anything done initially. It should also be noted that ensemble simulations are computationally more effective than long simulations. On a large supercomputer, all the sampling can be done concurrently obtaining the statistical results in the time it takes to run a single simulation. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat, or at least different starting velocities. Repeats or ensemble-based simulations will allow for a robust error statistic on single free energy calculations and are typically expected in the order of 1 kcal/mol or better.

@@

%A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}. The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties. To conclude, ensemble simulations can remove the aleatoric uncertainties in MD based predictions, however, further work is needed to ensure reproducibility across software packages arising due to systematic/parametric uncertainties.

@@

\subsection{Free energies of hydration or partition coefficients}
\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

\paragraph{The choice of solvent forcefield impacts the calculation}
A careful choice of forcefield for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of both water and the organic solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies, where solvent model parameters may vary substantially between force fields and impact results accordingly.

\paragraph{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as RESP or AM1-BCC, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{jakalian2000, basma2001, zhang2011, dupradeau2010, osato2025}.  Conformers with strong intramolecular interactions In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

\paragraph{Conformational sampling can impact solvation and partition free energies}
Moreover, while sampling challenges are often perceived to be minor for small molecules, they can still be non-trivial. Even subtle degrees of freedom such as hydroxyl group orientations (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. In some cases, enhanced sampling techniques such as replica exchange or metadynamics may be beneficial, even in hydration or partition coefficient calculations.

In partition coefficient calculations, the situation becomes even more complex, as two distinct environments must be adequately sampled: typically water and a nonpolar organic solvent. These systems not only differ in their dielectric properties and hydrogen bonding behavior but may also yield different dominant solute conformers. The partition free energy is sensitive to the relative population of conformers in each environment, so it is essential to ensure that sampling in each phase is independently converged. Some studies have employed restrained or confining potentials to prevent solutes from drifting to the interface (especially when using periodic systems with both phases present), while others model each phase separately and compute the free energy difference via thermodynamic cycles\cite{turchi2019,bannan2016,petris2021}.

\paragraph{Validation through experimental comparisons}  
Finally, experimental comparisons in these systems are often more direct than in binding free energies. Hydration and partition coefficients are frequently measured with high precision, enabling rigorous validation of computational protocols. This has led to their prominent use in blind challenges such as SAMPL, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies. These exercises have repeatedly shown that small changes in simulation parameters or protocols can result in significant deviations from experimental results, reinforcing the need for careful setup and validation even in seemingly "simple" solvation systems.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.

@@

\paragraph{Atom mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. \& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negligible due to the cancellation of errors. \cite{ries2024kartograf}

@@

A next step for atom mappings, would be to support multistate methods (e.g. RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2023fast}) with efficient mapping algorithms. There are protoype starting point for solving this problem, but to our knowledge, there is only few published work on this aspect. \cite{petrov2021perturbation}

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{ABFErest}). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, rieder2022lev}

\paragraph{Dummy atoms}

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.

@@

@@

>>>>

\newcommand{\versionnumber}{2.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/alchemistry/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[2]{Bryce K. Allen}
\author[11]{Agastya P. Bhati}
\author[18]{Stefan Boresch}
\author[3]{Hannah E. Bruce Macdonald}
\author[3*]{John D. Chodera}
\author[11, 12, 13]{Peter V. Coveney}
\author[15]{Jonathan W. Essex}
\author[19]{Emilio Gallicchio}
\author[9]{Vytautas Gapsys}
\author[23,24]{Michael Gillhofer}
\author[9]{David F. Hahn}
\author[20]{Wei-Tse Hsu}
\author[1,10]{Maximilian Kuhn}
\author[15]{Oliver J. Melling}
\author[1]{Julien Michel}
\author[4*]{David L. Mobley}
\author[5]{Levi N. Naden}
\author[23,24]{Chris Oostenbrink}
\author[4]{Meghan Osato}
\author[15,16]{William G. Poole}
\author[6]{Samarjeet Prasad}
\author[21,22]{Benjamin Ries}
\author[2,7]{Andrea Rizzi}
\author[17]{Jenke Scheen}
\author[14]{David R. Slochower}
\author[8*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[2]{Huafeng Xu}

%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Silicon Therapeutics, Boston, MA, USA}
\affil[3]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[4]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, Irvine, USA}
\affil[5]{Molecular Sciences Software Institute, Blacksburg VA, USA}
\affil[6]{National Institutes of Health, Bethesda, MD, USA}
\affil[7]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[8]{University of Colorado Boulder, Boulder, CO, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340, Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Centre for Computational Sciences, Department of Chemistry, University College London, London WC1H 0AJ, United Kingdom}
\affil[12]{Advanced Research Computing Centre, University College London, London WC1H 9BT, United Kingdom}
\affil[13]{Computational Science Laboratory, Institute for Informatics, Faculty of Science, University of Amsterdam, Amsterdam 1012, The Netherlands}
\affil[14]{Vertex Pharmaceuticals, San Diego, CA, USA}
\affil[15]{School of Chemistry and Chemical Engineering, University of Southampton, Southampton, SO17 1BJ, United Kingdom}
\affil[16]{Astex Pharmaceuticals, 436 Cambridge Science Park, Milton Road, Cambridge, CB4 0QA, United Kingdom}
\affil[17]{Open Molecular Software Foundation, Davis, CA 95618, USA}
\affil[18]{University of Vienna, Faculty of Chemistry, Institute of Computational Biological Chemistry, Wahringer Str. 17, A-1090 Vienna}
\affil[19]{Department of Chemistry and Biochemistry, Brooklyn College of the City University of New York, New York, NY, 11210
Ph.D. Program in Chemistry, The Graduate Center of the City University of New York, New York, NY, 10016
Ph.D. Program in Biochemistry, The Graduate Center of the City University of New York, New York, NY, 10016}
\affil[20]{Department of Biochemistry, University of Oxford, Oxford OX1 3QU, United Kingdom}
\affil[21]{Boehringer Ingelheim Pharma GmbH \& Co KG, Medicinal Chemistry, Birkendorfer Str 65, 88397 Biberach an der Riss, Germany}
\affil[22]{Open Free Energy, Open Molecular Software Foundation, Davis, CA, 95616, United State}
\affil[23]{Institute of Molecular Modeling and Simulation, BOKU University, 1190 Vienna, Austria}
\affil[24]{Christian Doppler Laboratory for Molecular Informatics in the Biosciences, BOKU University, 1190 Vienna, Austria}

\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Stefan Boresch}{0000-0002-2793-6656}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Emilio Gallicchio}{0000-0002-2606-4913}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{David R. Slochower}{0000-0003-3928-5050}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}
\orcid{David F. Hahn}{0000-0003-2830-6880}
\orcid{Vytautas Gapsys}{0000-0002-6761-7780}
\orcid{Agastya P. Bhati}{0000-0003-4539-4819}
\orcid{Peter V. Coveney}{0000-0002-8787-7256}
\orcid{William G. Poole}{0009-0003-2441-8794}
\orcid{Oliver J. Melling}{0000-0001-6243-7433}
\orcid{Jonathan W. Essex}{0000-0003-2639-2746}
\orcid{Wei-Tse Hsu}{0000-0001-6167-5480}
\orcid{Benjamin Ries}{0000-0002-0945-8304}
\orcid{Meghan Osato}{0000-0002-2732-457X}
\orcid{Michael Gillhofer}{0000-0001-8081-6373}
\orcid{Chris Oostenbrink}{0000-0002-4232-2556}

@@

MD trajectories are extremely sensitive to their starting conditions and hence one-off simulations can never be reproducible~\cite{coveney2016}. Thus, it is a necessity to perform ensemble simulations in order to perform ensemble averaging to obtain thermodynamic quantities in a reliable manner. Coveney and Wan recently published a book on the probabilistic foundations of MD and how this approach naturally allows for the treatment of uncertainties that arise in the use of the methodology~\cite{MDbook2025}. In practice, in many cases the simplest and most effective way of performing ensemble simulations is to vary only the initial velocities as these yield reproducible (i.e. statistically robust) free energy predictions. Bhati \textit{et al.} developed the TIES method~\cite{bhati2017, bhati2022, bhati2025} which is based on the use of ensemble methods. It should be noted that the size of ensemble, that is, the number of independent simulations, is an adjustable parameter whose optimal value is system-dependent and will also be affected by the desired level of precision. Further, this method is extremely flexible and adaptive allowing one to reduce the computational costs when resources are limited~\cite{wan2023}. Studies have shown that, in general, an ensemble size of 5 and 10 members is a reasonable starting point for alchemical RBFE and ABFE calculations, respectively, but recommendations are provided to adjust them based on some rules of thumb~\cite{bhati2022, bhati2025}. 
The sample standard deviation will give a crude estimate of the reliability of the estimates, and whether the precision is sufficient for the problem at hand. When practical, a more stringent test is to use different input coordinates for each repeat, or at least different starting velocities. Repeats or ensemble-based simulations will allow for a robust error statistic on single free energy calculations and are typically expected in the order of 1 kcal/mol or better.

@@

A recent study demonstrates reproducibility across software packages as well as free energy estimators, but only when ensemble simulations are employed~\cite{wade2022}. The authors reported results from two different MD engines (NAMD and OpenMM) as well as two free energy estimators (FEP and MBAR) for a set of 54 protein-ligand complexes which were found to be in good agreement when ensemble simulations were used to control uncertainties. To conclude, ensemble simulations can remove the aleatoric uncertainties in MD based predictions, however, further work is needed to ensure reproducibility across software packages arising due to systematic/parametric uncertainties.

@@

\subsection{Free energies of solvation or partition coefficients}
As described in Section~\ref{sec:theory-solvation}, the theory and equations for computing solvation free energies are well established. Here, we focus instead on practical considerations for applying these methods to hydration and partition coefficient calculations.

\label{subsec:hydration}
Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straightforward. For example, a 3D minimised structure of a solute can be generated with a simple tool such as openBabel and solvated to prepare the input to compute a free energy of hydration~\cite{oboyle2011open}. Despite their relative simplicity, care should still be taken when preparing these systems to avoid introducing systematic errors.

\paragraph{The choice of solvent force field impacts the calculation}
A careful choice of force field for the organic solvent model, as well as water model is essential. A critical aspect of these calculations is the preparation ahead of simulation. The choice of force field for both the solute and solvents involved is critical. Accurate modeling of the solute's solvation behavior hinges on realistic representations of the solvent (for partition coefficient calculations). See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion of these choices.  Beyond water, The same level of scrutiny is needed for organic solvents such as octanol, cyclohexane, or others used in partitioning studies. In a study by Open Force Field, they found that the GAFF\cite{wang2004gaff} and Sage \cite{boothroyd2023sage} forcefields produced results in non-aqueous solutions. Care should be taken when choosing the forcefield as solvent model parameters may vary substantially between force fields and impact results accordingly.

\paragraph{Charge models are non-trivial}
Modeling the partial charges of the solute and solvent molecules is also highly important. In some fixed-charge models, the assigned partial charges depend on the input conformation of the molecule, meaning that poor or inconsistent conformer selection can introduce systematic errors in solvation free energy estimates \cite{jakalian2000, jambeck2003}. This is particularly true for methods such as RESP\cite{bayly1993resp} or AM1-BCC\cite{jakalian2000}, where the geometry of the molecule at the time of charge derivation can influence the electrostatic potential fit \cite{basma2001, zhang2011, dupradeau2010, osato2025}. In these cases, the conformer(s) used for assigning the partial charges should be saved for reproducibility. Ensuring consistency between the charge derivation geometry and the initial simulation structure is therefore recommended, or alternatively, using ensemble-averaged charges or polarizable models where appropriate \cite{ren2002,lemkul2016}.

\paragraph{Equilibration of the Solvent}
Before computing solvation or partition free energies, it is important to ensure that the solvent itself is adequately equilibrated. While water typically equilibrates rapidly, especially when prepared using standard protocols, some organic solvents—particularly those with long hydrocarbon chains—can require substantially longer equilibration times\cite{fan2021sampl7}. Failing to properly equilibrate the solvent can introduce systematic errors in free energy calculations, as the solvent structure and density may not reflect the true thermodynamic ensemble of the target environment\cite{zhang2017solventmodel}.

\paragraph{Conformational sampling can impact solvation and partition free energies}
Moreover, while sampling challenges are often perceived to be minor for small molecules, they can still be non-trivial. Even subtle degrees of freedom such as hydroxyl group orientations (as in neutral carboxylic acids) or intramolecular hydrogen bonding can influence the calculated free energy difference between two environments\cite{klimovich2010predicting, lim2019assessing}. Molecules with multiple rotatable bonds, internal hydrogen bonds, or conformational flexibility may exhibit environment-dependent populations that are difficult to sample adequately within the timescale of typical molecular dynamics simulations. Typically, solvation free energies are computed as absolute transformations, meaning that as the size and flexibility of the ligand increase, the sampling burden grows and convergence of the free energy estimate becomes more challenging. In some cases, enhanced sampling techniques such as replica exchange\cite{rest2} or metadynamics\cite{hsu2023alchemical, khuttan2024make} may be beneficial, even in hydration or partition coefficient calculations.

In partition coefficient calculations, the situation becomes even more complex, as two distinct environments must be adequately sampled: typically water and a nonpolar organic solvent. These systems not only differ in their dielectric properties and hydrogen bonding behavior but may also yield different dominant solute conformers. The partition free energy is sensitive to the relative population of conformers in each environment, so it is essential to ensure that sampling in each phase is independently converged. It is also important to ensure adequate overlap in the common end states—if the solute is collapsed to two different gas-phase or non-interacting states, the thermodynamic cycle may not hold\cite{fan2021sampl7}. Some studies have employed restrained or confining potentials to prevent solutes from drifting to the interface (especially when using periodic systems with both phases present), while others model each phase separately and compute the free energy difference via thermodynamic cycles\cite{turchi2019,bannan2016,petris2021}.

\paragraph{Validation through experimental comparisons}  
Finally, experimental comparisons in these systems are often more direct than in binding free energies. Hydration and partition coefficients are frequently measured with high precision, enabling rigorous validation of computational protocols. Calculating solvation free energies for charged solutes, however, remains particularly challenging, as long-range electrostatics, finite-size artifacts, and treatment of counterions can significantly affect results. Moreover, direct experimental determination of ionic hydration free energies is generally not feasible; instead, such values are inferred by decomposing the solvation free energy of ion pairs \cite{mobley2014freesolv}. This has led to their prominent use in blind challenges such as SAMPL \cite{skillman2012sampl3, mobley2014sampl4, bergazin2021sampl7}, where participants predict hydration or partition free energies of small molecules using a variety of force fields, solvent models, and sampling strategies. These exercises have repeatedly shown that small changes in simulation parameters or protocols can result in significant deviations from experimental results, reinforcing the need for careful setup and validation even in seemingly "simple" solvation systems.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s middle branch depicts how such a dual topology works. Note that the two groups that are switched on or off never see each other. The 'mixed' intermediate atoms (M) are used as well, but they strictly interpolate between physical and dummy atoms, or vice versa.
Bhati \textit{et al.}~\cite{bhati2017, bieniek2021} presented a variant of the single coordinate dual topology approach which is used in the TIES method. Its distinguishing feature is the application of some filters to remove atoms with differences in atom types and partial charges from the structural MCSS identified. This is followed by ensuring identical charges for all MCSS atoms. The approach ensures that the final MCSS is both structurally and chemically identical, allowing the use of a unique set of MCSS atoms for both ligands and dual topology for their alchemical parts.

@@

\paragraph{Atom mapping} \label{subsec: Atom Mapping}
Following the choice of system representation and sampling strategy, it might be required to generate a atom mapping. This is the case especially for system representations using a single coordinate set (Type 1. \& 2), however a dual coordinate set approach could use an atom mapping as well, depending on the sampling method, this will be explained in the next steps.
Once a single coordinate set approach is selected, a crucial next step is to identify the common atoms which will not be perturbed between the two molecules (respectively end-states).
Rigorously, this process typically comprises a MCSS search of the molecules involved to identify the common substructure.
Specifically, with a single coordinate set approach in mind, atom types are allowed to change, so a permissive MCSS search can be done. This however requires either careful implementation of atom mass changes, bond length etc. or accepting the introduction of potential systematic error, that is in many cases negligible due to the cancellation of errors. \cite{ries2024kartograf}

@@

A next step for atom mappings, would be to support multistate methods (e.g. RE-EDS~\cite{sidler2016replica}, A-EDS~\cite{perthold2020toward} or Ladybugs~\cite{robo2023fast}) with efficient mapping algorithms. There are starting points, suggesting protoype algorithms for solving this problem, but to our knowledge, there is only few published work on this aspect. \cite{petrov2021perturbation}

Single coordinate set relative calculations and calculations based on substructure searches only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}.
If no common substructure is shared, then dual coordinate sets methods like linked-dual or separated topology free energy calculations are needed. One would co-localize a pair of compounds in a binding site, exclude their interactions with one another, and compute the relative binding free energy by turning one molecule on from being dummy atoms while turning the other off.
For such approaches it is crucial to find a good set of restraints, that keep the ligands colocalized over the time of the simulation . This problem is similar to the required restraints in ABFE Calculations (see \ref{sec:ABFE-standard-state} ). Few descriptions of possible approaches in RBFE calculations can be found for such approaches like the one suggested by \textit{Baumann et al.}~\cite{baumann2023broadening} based on Boresch style restraints~\cite{boresch2003absolute}, with an elaborate automatic atom selection approach, for the seperated topology approach, or Restraintmaker~\cite{ries2022restraintmaker} for the linked-DCDT approaches, which finds a small set of atoms to be linked via distance restraints, with a small distance to each other. 
However the implementation and testing of an fully automatized dual-topology pipeline for RBFE dual topology approaches is still to our knowledge a remaining research problem and was only tested with small datasets. \cite{jespers2019qligfep, rieder2022lev}

\paragraph{Dummy atoms} \label{subsec: Dummy Atoms}

@@

One situation involving dummy atoms that needs to be avoided is the creation of a cyclic structure involving dummy atoms, i.e., whenever the (group of) dummy atom(s) is attached to two different parts of the physical molecule. In this case, the factorization of the partition function is never possible \cite{Shobana_2000}. It follows that ring opening or closing is extremely challenging to carry out using SCST approaches \cite{liu2015ring}; these difficulties have in part motivated the development and adoption of dual coordinate approaches. For more on ring opening/closing, see also below. Nevertheless, DC approaches are susceptible to exactly this type of problem: to keep the noninteracting molecule on top of the interacting one, typically multiple bonded restraints (e.g. harmonic bonds) are used \cite{Axelsen_1998,Ries_2022}. These effectively, are equivalent to the formation of cyclic structures with potential erroneous contributions to the double free energy difference of interest. The type of restraints used in SepTop \cite{rocklin2013separated,Baumann_2023} and ATM \cite{Azimi_2022} avoid this problem.
Additionally, TIES~\cite{bhati2017} circumvents this issue by design as it does not need any restraints.

@@

\textbf{MG} Contributed to ``Changes in net charge can be challenging/problematic'' in Sec. \ref{subsec:net charge}. \\

@@

\textbf{MO} Contributed to Section \ref{subsec:hydration}
